\section{Introduction}

%\subsection{Overview}

Functional programming languages, like ML and Haskell,
come with strong static types system, which detects a lot 
of errors at compile-time and enhances code documentation.


The utility of these type systems stems from their ability to predict, 
at compile-time, invariants about the run-time values computed by the program. 
%
Unfortunately, classical type systems only
capture relatively coarse invariants. For example, the system can
express the fact that a variable @i@ is of the type @Int@, 
meaning that it is always an integer, 
but not that it is always an integer with a
certain property, say different than zero. 
%
Thus, the type system is unable to statically ensure the safety of critical operations, 
such as a division by i. 
%
Several authors have proposed the
use of refinement types \cite{FreemanPfenning91, flanagan06, LiquidPLDI08, Greenberg12} as a mechanism for enhancing the
expressiveness of type systems. 

\textit{Refinement types} refine a vanilla type with a predicate.
For example, one can give @i@ the following type:
$$\centering\begin{tabular}{c}
\begin{code}
i :: {v:Int | v != 0}
\end{code}
\end{tabular}$$

that describes a value @v@ of type @Int@,
while, the refinement constraints 
this value @v@ to be different than @0@.


One can use this refinement type to define a safe division operator:
%
$$\centering\begin{tabular}{c}
\begin{code}
safeDiv :: Int -> {v:Int | v != 0} -> Int
\end{code}
\end{tabular}$$
%
This type captures that the division operator takes two
@Int@ arguments and returns an @Int@.
Moreover, it restricts the second argument to be different that
zero, to eliminate division by zero operations. 

At the call side of @safeDiv@ 
the type system should check that the real arguments do not
violate its specification.
%
For instance, 
@safeDiv 8 9@ is safe, since @9@ is always different that zero.
But, @safeDiv 8 0@ should create a type error.
Apart from concrete values, @safeDiv@ can be applied to arbitrary program expressions:
@safeDiv n m@ is safe only if 
@m@ is an integer different than zero.

\mypara{Refinement Function Types}
Refinement function types\cite{cayenne, flanagan06}, allow the specification of the 
result to depend on the argument.
%
A parameter is used to bind the argument
and can appear in the refinement of the result.
%
%Dependent types, allow the refinement of the result of a function
%to depend on function's arguments.
As an example, we define a @pred@ function, 
that takes as argument a positive integer @n@ and returns the its predecessor.
Refinement function types allow us to give @pred@ a type
that exactly captures this behaviour:
%
$$\centering\begin{tabular}{c}\begin{code}
pred :: n : {v:Int | v > 0} -> {v:Int | v = n-1}
pred n = n-1
\end{code}\end{tabular}$$
%
This type expresses that for each positive integer argument @n@, 
the result is an @Int@ exactly equal to @n-1@.
When @pred@ is applied to a concrete value, 
the parameter @n@ is substituted with this value.
For example @pred 2 :: {v:Int | v = n-1}[2/n] = {v:Int | v = 1}@.
Thus, for each concrete argument, the result should be the predecessor 
of this argument.

\mypara{Verification}
\textit{Verification} is a procedure that takes as input 
a program, i.e.,  definitions for functions and values,
and some specifications, i.e., refinement type signatures for functions and values,
and decides whether the specifications hold for the program.
%
Informally, it checks that 
each expression satisfies its type, 
for example, the @pred@ definition actually returns the 
predecessor of its argument,
or that 
at each function application the arguments satisfy
the function's preconditions, 
as in the @safeDiv@ example.
%
If the specifications hold, the program is \textit{Safe}, 
otherwise it is \textit{Unsafe}. 

Higher-order programming languages, such as ML or Haskell, 
treat functions as first order objects.
Thus, one can use functions in refinements and 
create higher-order predicates. 
For instance, the following type
%
$$\centering\begin{tabular}{c}\begin{code}
f:(a->b) -> {v:Bool|terminates f}
\end{code}\end{tabular}$$
%
describes that an arbitrary functional argument
should satisfy a predicate @terminates@.
%
Reasoning in a higher-order logic is undecidable, 
thus if arbitrary program values appear in the refinements,
the verification procedure is undecidable.
%
As we shall see, 
if the refinement language is restricted, i.e., is less expressive,
verification can be decidable.

The rest of this paper is organized as follows:
In section \ref{subsec:formal}, we present a core calculus
that constitutes the base for many refinement type systems.
In section \ref{sec:undec},
we describe reasoning in undecidable refinement type systems.
In section \ref{sec:liquid}, 
we present less expressive type systems
that are decidable.
In section \ref{sec:abstract} 
we present, how abstraction over refinements
enhances expressiveness of decidable type systems. 
Finally, we conclude.

\section{Preliminaries}\label{subsec:formal}

To formally describe and compare type systems, 
we define a core calculus, following \cite{flanagan06, LiquidPLDI08, Greenberg12}.
We refer to our calculus as $\lambda_C$, 
and in this section we present its syntax and type system.

\subsection{Syntax}
The syntax of expressions and types is summarized in Figure \ref{fig:coresyntax}.
$\lambda_C$ expressions include variables, constants,
typed $\lambda$-abstractions
and function applications. 
%
Constants include primitive integers:
$0, 1, 2, \dots$ and primitive booleans: \texttt{true} or \texttt{false},
which take the basic types, integer and boolean, respectively.
%
A basic type can be refined with a predicate to construct a basic
refinement type.
Refinement types also contain function types, 
in which, 
a variable binds the argument,
so that the result refinement can refer to it.  
%
The predicate $p$ is
not yet defined. 
As noted earlier, if $p$ contains arbitrary program 
expressions, the type system is undecidable, but 
$p$ can be restricted in such a way as to render 
the type system decidable.
%
Finally, we define a typing environment $\Gamma$, that maps variables to their type,
and will be used in the rules.


\begin{figure}[t!]
\centering
$$
\begin{array}{rrcl}
\emphbf{Expressions} \quad 
  & e 
  & ::= 
  &      x 
  \spmid c 
  \spmid \efunt{x}{\tau}{e} 
  \spmid \eapp{e}{e} 
%  \spmid \etabs{\alpha}{e} 
%  \spmid \etapp{e}{\tau} 
  \\[0.05in] 

\emphbf{Predicates} \quad 
  & p
  & ::= 
  & \dots
  \\[0.05in] 

\emphbf{Basic Types} \quad 
  & b 
  & ::= 
  &      \tbint
  \spmid \tbbool
%  \spmid \alpha
  \\[0.05in]

\emphbf{Refinement Types} \quad 
  & \tau 
  & ::= 
  &      \tref{b}{p} 
  \spmid \tfun{x}{\tau}{\tau}
  \\[0.05in]

\emphbf{Typing Environment} \quad 
  & \Gamma 
  & ::= 
  &      \emptyset 
  \spmid x:\tau, \Gamma

\end{array}
$$
\caption{\textbf{Syntax of $\lambda_C$}}
\label{fig:coresyntax}
\end{figure}



\subsection{Typing}
The typing rules used by $\lambda_C$ are summarized in Figure \ref{fig:corerules}.

% \subsubsection*{typing}
\mypara{Type Checking}
Type checking rules state that
an expression $e$ has a type $\tau$ under an environment $\Gamma$,
that is, when the free variables in $e$ are bound to values described by 
$\Gamma$, the expression $e$ will evaluate to a value described by $\tau$.
We write \hastype{\Gamma}{e}{\tau} and create one rule for each program expression.

The rule \tconst uses a function $tc$ that maps each primitive constant 
to its predefined type.
%
The rule \tvariable checks the type of a variable, according 
to the environment $\Gamma$.
%
The rule \tfunction checks the type of the function-body
in the environment, extended with the argument of the function.
Since the argument type is given, it could be any arbitrary type,
say \tref{b}{1},
which is invalid, as a base type is refined with 
the value $1$, which can not be a valid predicate.
%
A wellformedness rule is used to 
check that the argument type is \textit{well-formed}, 
i.e., its refinements are valid predicate expressions.
%
Finally, the rule \tapp checks that in an application \eapp{e_1}{e_2}
the expression $e_1$ has a function type
whose argument is the type of the argument $e_2$.
%
As we discussed in the introduction, in the final type,
the parameter $x$ should be replaced with the actual argument $e_2$.

%\subsubsection{wf}
\mypara{Wellformedness rules}
Wellformedness rules  
state that a type $\tau$ is well-formed under environment
$\Gamma$, that is, the refinements in $\tau$ are boolean 
expressions in the environment $\Gamma$.
%
We write \isWellFormed{\Gamma}{\tau} and create one rule for each type.

The rule \wtBase checks that in a basic refinement type,  
the refinement is a valid boolean expression.
The environment of this check is extended with the value that is refined;
for instance, to check the validity of \tref{\texttt{int}}{v > 0}, 
we check that $v > 0$ is a boolean expression, in an 
environment where $v$ is an integer value.
%
The rule \wtFun recursively applies the wellformedness rule to
the argument type of the function, and to the result type, 
in an environment extended with the argument parameter.

%\subsubsection{sub}
\mypara{Subtyping rules}
Consider that the predefined type for the integer $2$
is an integer that is exactly equal to $2$.
The type system can check,via the rule \tconst , that 
\hastype{\emptyset}{2}{\tref{\texttt{int}}{v = 2}}.
If $2$ is applied to a function that expects a
positive integer, say $f::\tfun{x}{\tref{\texttt{int}}{v>0}}{\tau}$,
the type system should also check that 
\hastype{\emptyset}{2}{\tref{\texttt{int}}{v > 0}}.
%
There are many ways for this check to succeed.
We will present \textit{syntactic subtyping},
in which subtyping reduces to implication checking.
In our example, $v = 2 \Rightarrow v > 0 $ implies 
$\tref{\texttt{int}}{v=2} \preceq \tref{\texttt{int}}{v>0}$.


In the general case, subtyping rules
state that the type $\tau_1$ is a subtype of the type
$\tau_2$ under environment $\Gamma$, that is, when the free variables
of $\tau_1$ and $\tau_2$
are bound to values described by $\Gamma$, the set of values described
by $\tau_1$ is contained in the set of values described by $\tau_2$. 
We write \isSubType{\Gamma}{\tau_1}{\tau_2} and create one rule for every type:

The rule \tsubBase serves two purposes:
Firstly,
it checks that the basic type is the same in the two types.
Moreover, it checks that under the environment $\Gamma$, 
the left hand side refinement implies the right hand side.
The implpication checking is enforced by a predicate \texttt{Valid} 
which varies between the systems that we will describe.
%
The rule \tsubFun relates two function types according to the contravariant rule.


In the rest of this paper, we will the core calculus $\lambda_C$
upon which we will build a subset of four typing systems\cite{TODO}.


\begin{figure}[ht!]
\medskip \judgementHead{Type Checking}{$\hastype{\Gamma}{e}{\tau}$}

\begin{comment}
$$\begin{array}{cc}

\inference
  {  \hastype{\Gamma}{e}{\tau_2} && \isSubType{\Gamma}{\tau_2}{\tau_1} 
  && \isWellFormed{\Gamma}{\tau_1}
  }
  {\hastype{\Gamma}{e}{\tau_1}}
  [\tsub]
\end{array}$$
\end{comment}
$$\begin{array}{cc}

\inference
  {}
  {\hastype{\Gamma}{c}{\tc{c}}}
  [\tconst]
&
\inference
  {x:\tau \in \Gamma}
  {\hastype{\Gamma}{x}{\tau}} 
  [\tvariable]

\end{array}$$

$$\begin{array}{cc}

\inference
  {\hastype{\Gamma, x:\tau_x}{e}{\tau} &&     
  \isWellFormed{\Gamma}{\tau_x}
  }
  {\hastype{\Gamma}{\efunt{x}{\tau_x}{e}}{\tfun{x}{\tau_x}{\tau}}}
  [\tfunction]
&
\inference
  {\hastype{\Gamma}{e_1}{\tfun{x}{\tau_x}{\tau}} &&
   \hastype{\Gamma}{e_2}{\tau_x}
  }
  {\hastype{\Gamma}{\eapp{e_1}{e_2}}{\tau[e_2/x]}} 
  [\tapp]

\end{array}$$

\judgementHead{Well-Formedness}{\isWellFormed{\Gamma}{\tau}}

$$\inference
    {\hastype{\Gamma, v:b}{p}{\tbbool}}
    {\isWellFormed{\Gamma}{\tref{b}{p}}}
    [\wtBase]
\qquad
\inference
    {
    \isWellFormed{\Gamma}{\tau_x} &&
	\isWellFormed{\Gamma, x:\tau_x}{\tau}
    }
    {\isWellFormed{\Gamma}{\tfun{x}{\tau_x}{\tau}}}
    [\wtFun]
$$


\medskip \judgementHead{Subtyping}{\isSubType{\Gamma}{\tau}{\tau}}

$$
\inference
   {(\Gamma, v:b \vdash \texttt{Valid}(p_1 \Rightarrow p_2))}
   {\isSubType{\Gamma}{\tref{b}{p_1}}{\tref{b}{p_2}}}
   [\tsubBase]
$$

$$
\inference
   {\isSubType{\Gamma}{\tau_{21}}{\tau_{11}} &
	\isSubType{\Gamma, x_2:{\tau_{21}}}{\SUBST{\tau_{12}}{x_1}{x_2}}{\tau_{22}}	
   }
   {\isSubType{\Gamma}
	  {\tfun{x_1}{\tau_{11}}{\tau_{12}}}
	  {\tfun{x_2}{\tau_{21}}{\tau_{22}}}
}[\tsubFun]
$$
\caption{\textbf{Static Semantics for $\lambda_C$}}
\label{fig:corerules}
\end{figure}